---
title: "HW2"
output: github_document
date: "2022-09-29"
---

# Problem 1

Before we do anything, load the `tidyverse` package.

```{r, message = FALSE}
library(tidyverse)
```

Now we can get started!

First, load the data.
```{r, message = FALSE}
# Read data and clean names
transit = read_csv("NYC_Transit_Subway_Entrance_And_Exit_Data.csv", 
                   col_types = cols(Route8 = "c", Route9 = "c", Route10 = "c", Route11 = "c"))  %>% 
  janitor::clean_names()
  
# Check column names of imported data
colnames(transit)

```

Next we do some cleaning.
```{r}
transit_clean = transit %>% 
 select(line:entry, vending, ada) %>% 
  mutate(entry = ifelse(entry == "YES", TRUE, FALSE))

```

transit_clean is a `r nrow(transit_clean)`x`r ncol(transit_clean)` dataset with the following variables:

* ``r colnames(transit_clean)``

However, there are still many separate variables in transit_clean that could be combined to be tidier.

Here are some other details about the dataset.

```{r}
# Number of distinct stations
nrow(distinct(transit_clean, station_name))
```

There are **`r nrow(distinct(transit_clean, station_name))`** distinct stations.

```{r}
# ADA-compliant stations

transit_clean %>% filter(ada == TRUE) %>% nrow()

```
**`r transit_clean %>% filter(ada == TRUE) %>% nrow()`** stations are ADA-compliant.

```{r}
# Proportion of station entrances/exits without vending that allow entry
transit_clean %>% 
  filter(vending == "NO") %>% 
  pull(entry) %>% 
  mean()

```
**`r transit_clean %>% filter(vending == "NO") %>% pull(entry) %>% mean()*100`%** of station entrances/exits without vending allow entry.


Let's make transit_clean tidy by reformatting the route number and name with `pivot_longer`, and then filter by distinct stations that serve the A train and how many of those are ADA compliant.

```{r}
transit_tidy = 
  transit_clean %>% 
  pivot_longer(route1:route11, 
               names_to = "route_number",
               values_to = "route_name") 

transit_tidy %>% 
  filter(route_name == "A") %>% 
  select(station_name, line) %>% 
  distinct

transit_tidy %>% 
  filter(route_name == "A", ada == "TRUE") %>% 
  select(station_name, line) %>% 
  distinct
```
**`r transit_tidy %>% filter(route_name == "A") %>% select(station_name, line) %>% distinct %>% count()`** distinct stations serve the A train, and **`r transit_tidy %>% filter(route_name == "A", ada == "TRUE") %>% select(station_name, line) %>% distinct %>% count()`** distinct stations serve the A train and are ADA compliant.

# Problem 2

First, load the additional required packages for this problem.
```{r, message = FALSE}
library(readxl)
```



Now, I can load the Mr. Trash Wheel sheet within the Trash Wheel Collection Excel file. I specified a range that excludes the figures and columns containing notes, as well as omitting the final row indicating a grand total. 

After cleaning the variable names with `clean_names`, I renamed the weight_tons and volume_cubic_yards variables to be shorter. 
I filtered away observations that did not include dumpster-specific data with `filter`, and used `mutate` to round the number of sports balls to the nearest integer, and add a variable indicating this dataset was from the Mr. Trash Wheel sheet.
```{r}
mrtrash = read_excel("./data/Trash Wheel Collection Data.xlsx", 
                   sheet = "Mr. Trash Wheel", 
                   range = "A2:N549") %>% 
  janitor::clean_names() %>% 
  rename(weight = weight_tons,
                volume = volume_cubic_yards) %>% 
  filter(!is.na(dumpster)) %>% 
  mutate(sports_balls = as.integer(sports_balls),
         year = as.double(year),
         sheet = "Mr. Trash Wheel")
```

Then I did the same with the Professor Trash Wheel sheet:

```{r}
proftrash = read_excel("./data/Trash Wheel Collection Data.xlsx", 
                   sheet = "Professor Trash Wheel", 
                   range = "A2:M96") %>% 
  janitor::clean_names() %>% 
  rename(weight = weight_tons,
                volume = volume_cubic_yards) %>% 
  filter(!is.na(dumpster)) %>% 
  mutate(sheet = "Professor Trash Wheel")
```

Now I can use a `full_join` to combine the datasets into one dataset called trash.
```{r, message = FALSE}
trash = full_join(mrtrash, proftrash)
```

The combined data set consists of **`r nrow(trash)` observations and `r ncol(trash)` variables**.
The included variables are *`r colnames(trash)`*, with the last variable being an indicator of the original sheet the observation was taken from.

The total weight of trash collected by Professor Trash Wheel was **`r sum(proftrash$weight)` tons**.

The total number of sports balls collected by Mr. Trash Wheel in 2020 was **`r mrtrash %>% filter(year == 2020) %>% pull(sports_balls) %>% sum()` tons**.


# Problem 3

First, I loaded the pols dataset, then used `separate()` to break up the mon variable into year, month and day. Then I used `lubridate::month` to replace the month number with the month name. I then used `pivot_longer` to create a president variable taking values from prez_dem and prez_gop. Finally, I removed the day variable with `select`.

```{r, message = FALSE}
pols = read_csv("./data/pols-month.csv") %>% 
  separate(mon, into = c("year", "month", "day"), sep = "-") %>% 
  mutate(month = lubridate::month(as.numeric(month), label = TRUE, abbr = FALSE),
         year = as.numeric(year),
         president = ifelse(prez_gop == 1, "gop", "dem")) %>% 
  select(-day, -prez_gop, -prez_dem)

```


Second, clean the data in snp.csv using a similar process to the above. For consistency across datasets, arrange according to year and month, and organize so that year and month are the leading columns.
I cleaned the snp dataset in a similar process as above. Since the date variable was in a mm/dd/yy format, I first used `lubridate::mdy` to reformat it into the standard yyyy-mm-dd format, then I separated it into year, month and day variables with `select`, and changed the month into the full name and removed the day variable.

```{r, message = FALSE}

snp = read_csv("./data/snp.csv") %>% 
   mutate(date = lubridate::mdy(date)) %>% 
  separate(date, into = c("year", "month", "day"), sep = "-") %>% 
  mutate(month = lubridate::month(as.numeric(month), label = TRUE, abbr = FALSE),
         year = as.numeric(year)) %>% 
 select(-day)

  

```


Next, I tidied the unemployment data by using `pivot_longer` to create a single month variable instead of having one variable per month, then cleaned the names with `janitor::clean_names()` to be consistent with the previous datasets. I changed the month variable from the abbreviated month names to the full month names by using `factor` with the abbreviations (month.abb) as the levels and renaming the labels to the full month names (month.name).
```{r, message = FALSE}
unemployment = read_csv("./data/unemployment.csv") %>% 
  pivot_longer(Jan:Dec,
               names_to = "month",
               values_to = "unemployment") %>% 
  janitor::clean_names() %>% 
  mutate(month = factor(month, levels = month.abb, labels = month.name, ordered = TRUE),
         year = as.numeric(year))
```


Join the datasets by merging snp into pols, and merging unemployment into the result.

Finally, I used `left_join` to merge snp into pols, and then unemployment into the earlier merge.
```{r}
merge_df = left_join(pols, snp) %>% 
  left_join(., unemployment)
```


The pols dataset contained **`r nrow(pols)`** observations and **`r ncol(pols)`** variables. The included variables were *`r colnames(pols)`*.

The snp dataset contained **`r nrow(snp)`** observations and **`r ncol(snp)`** variables. The included variables were *`r colnames(snp)`*.

The unemployment dataset contained **`r nrow(unemployment)`** observations and **`r ncol(unemployment)`** variables. The included variables were *`r colnames(unemployment)`*.

The resulting merge_df dataset has a dimension of **`r nrow(merge_df)` observations x `r ncol(merge_df)` variables**. The range of years was `r range(merge_df$year)`, and the variable names are *`r colnames(merge_df)`*. 









